Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-25 src/1.Introduction.tex
Fix a few typos

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namelyexecution patterns and data structures which it maps to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

>>>>

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely execution patterns and data structures, which it maps to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at Sandia National Laboratories, University of Utah, Oak Ridge National Laboratory, Los Alamos National Laboratory and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

### 2020-01-25 src/4.ProgammingModel.tex
Fix a few typos

The Kokkos programming model specifies a language, programming paradigm and an API. The API exposes the machine representation to the developer and defines a set of pattens and types that capture intents and properties of execution. This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. This allows the programming model to  optimize load and store operations.
\end{itemize}

>>>>

The Kokkos programming model specifies a programming paradigm and an API. The API exposes the machine representation to the developer and defines a set of patterns and types that capture intents and properties of execution. This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. This allows the programming model to  optimize load and store operations.
\end{itemize}

### 2020-01-25 src/5.Example.tex
Fix a few typos

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications host-accessible allocation is necessary in order to allow I/O operations during fileystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

@@

Taking a closer look at the loop step shows the hierarchical expression of concurrency. In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}. For each row of~\emph{M} a team of threads is launched, where the size of the team is not further specified. When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads. 
This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies. The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index. It then writes the result to the correct position of the output vector. 
The parallel loops are follows by a call to \emph{Kokkos::fence}. This ensures that the execution of the parallel code is completed before copying the results back to the host memory space. This is implemented through the~\emph{Kokkos::deep\_copy} call. Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.

>>>>

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications host-accessible allocation is necessary in order to allow I/O operations during filesystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

@@

Taking a closer look at the loop step shows the hierarchical expression of concurrency. In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}. For each row of~\emph{A} a team of threads is launched, where the size of the team is not further specified. When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads. 
This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies. The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index. It then writes the result to the correct position of the output vector. 
The parallel loops are follows by a call to \emph{Kokkos::fence}. This ensures that the execution of the parallel code is completed before copying the results back to the host memory space. This is implemented through the~\emph{Kokkos::deep\_copy} call. Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.
